---
title: 'Physical activity and mental health in Canadian seniors: <br> A cross-sectional
  study using the Canadian Community Health Survey'
author: 
date:
output:
  html_document: 
    toc: yes
    toc_depth: 5
    toc_float:
      collapsed: no
      smooth_scroll: yes
  pdf_document: default
---

<style type="text/css">

body{ /* Normal  */
      font-size: 16px;
  }
td {  /* Table  */
  font-size: 16px;
}
h1.title {
  font-size: 26px;
  color: Black;
}
h1 { /* Header 1 */
  font-size: 22px;
  color: MidnightBlue;
}
h2 { /* Header 2 */
    font-size: 18px;
  color: MediumBlue;
}
h3 { /* Header 3 */
  font-size: 16px;
  color: DodgerBlue;
}
h4 { /* Header 4 */
  font-size: 15px;
  color: DeepSkyBlue;
}
code.r{ /* Code block */
    font-size: 12px;
}
pre { /* Code block - determines code spacing between lines */
    font-size: 14px;
}
</style>



**Authors:** *Reza Hosseini* and *Alexandra Roine*, MPH students at UBC  
**December 8, 2021**  

\

***
# SYNOPSIS
The Canadian Physical Activity Guidelines (CPAG) suggest that seniors participate 
in 150 minutes of moderate to vigorous exercise each week; however, Canadian 
seniors have not been meeting these guidelines. Exercise improves physical and 
mental health, and maintenance of these throughout ageing is important. The 
impact of exercise participation on the mental health of Canadian seniors 
remains unknown. This study aimed to investigate the association between 
exercise participation and mental health in Canadian seniors using data from 
the 2017-2018 Canadian Community Health Survey (CCHS). Bivariate and logistic 
regression models were used to investigate the effect of exercise participation 
on perceived mental health. CCHS respondents over 60 years old (N=25,806) were 
included for analysis. Respondents physically active below CPAG were older, and 
less likely to be male, injured, or employed than respondents who met CPAG. The 
unadjusted odds of a respondent who met CPAG reporting Excellent-Good mental 
health were 1.38 (95% CI: 1.24, 1.54) times that of a respondent below CPAG. 
After adjusting for age, sex, race, injury, and employment, the odds ratio was 
1.56 (95% CI: 1.37, 1.78). Sensitivity analysis showed missingness did not 
impact effect measures meaningfully. Employment status was an effect modifier, 
and the adjusted odds ratios for employed and unemployed respondents were 
1.09 (95% CI: 0.85, 1.40) and 1.81 (95% CI: 1.55, 2.12), respectively. Findings 
suggest that exercise participation at CPAG levels by Canadians aged 60 years 
and above may improve perceived mental health.

\

***
# LOAD LIBs & DATA

## Libraries
```{r, message=FALSE, warning=FALSE}
library("checkpoint") 
checkpoint("2021-01-01")  # checkpoint used to ensure reproducibility

library("dplyr") 
library("ggplot2") 
library("tidyr")
library("forcats")
library("tableone")       # Used for creating table one
library("kableExtra")     # Used to present table one in markdown
library("Publish")        # Used for displaying regression outputs
```


## Data

You can download the CCHS 2017-18 dataset 
[**here**](https://github.com/rezahosseinimed/portfolio2_CCHS/raw/main/cchs1718_original.RData.zip).  

It has `113,290` observations (respondents) and `1051` variables:

```{r}
load("cchs1718_original.RData")

dim(cchs1718_original)
```


\

***
# CREATE ANALYT. DATASET

## Selecting variables
Only 7 out of 1051 variables from the original CCHS dataset are selected.

```{r}
cchs_withExclusions <- cchs1718_original %>% 
  select(GEN_015,  # Outcome: Perceived mental health
         PAADVACV, # Exposure: Physical activity indicator
         DHHGAGE,  # Age
         DHH_SEX,  # Sex
         SDCDGCGT, # Race
         INJDVSTT, # Injuries
         MAC_010,  # Employment status
         )
```


## Redefining levels of variables
```{r}
cchs_withExclusions <- cchs_withExclusions %>%
  
  # Outcome: Perceived mental health
  mutate(pMentalHealth = case_when(GEN_015 %in% c("Fair",
                                                  "Poor") ~ "Fair-Poor",
                                   GEN_015 %in% c("Excellent",
                                                  "Very good",
                                                  "Good") ~ "Excellent-Good")) %>%
  
  # Explanatory variable: Physical activity indicator
  mutate(physicalAct = case_when(
    PAADVACV == paste0("Physically active below ",
                       "recommended level from CPAG") ~ "below act. level",
    PAADVACV == paste0("Physically active at ",
                       "/ above recommended ",
                       "level from CPAG") ~ "above act. level")) %>% 
  
  # Covariate: Age
  mutate(age = case_when(DHHGAGE == "Age between 12 and 14" ~ "12-14",
                         DHHGAGE == "Age between 15 and 17" ~ "15-17",
                         DHHGAGE == "Age between 18 and 19" ~ "18-19",
                         DHHGAGE == "Age between 20 and 24" ~ "20-24",
                         DHHGAGE == "Age between 25 and 29" ~ "25-29",
                         DHHGAGE == "Age between 30 and 34" ~ "30-34",
                         DHHGAGE == "Age between 35 and 39" ~ "35-39",
                         DHHGAGE == "Age between 40 and 44" ~ "40-44",
                         DHHGAGE == "Age between 45 and 49" ~ "45-49",
                         DHHGAGE == "Age between 50 and 54" ~ "50-54",
                         DHHGAGE == "Age between 55 and 59" ~ "55-59",
                         DHHGAGE == "Age between 60 and 64" ~ "60-64",
                         DHHGAGE == "Age between 65 and 69" ~ "65-69",
                         DHHGAGE == "Age between 70 and 74" ~ "70-74",
                         DHHGAGE == "Age between 75 and 79" ~ "75-79",
                         DHHGAGE == "Age 80 and older" ~ "80+")) %>%
  
  # Covariate: Sex
  rename(sex = DHH_SEX) %>%
  
  # Covariate: Race
  mutate(race = case_when(
    SDCDGCGT == "White" ~ "White",
    SDCDGCGT == "Non-white (Aboriginal or Other Visible Minority)" ~ "Non-white")) %>%
  
  # Covariate: Injuries:
  mutate(injury = case_when(
    INJDVSTT == "No injuries" ~ "No injuries",
    INJDVSTT == "Injury limiting activities only" ~ "Limiting activities only",
    INJDVSTT == "Treated injury (not limiting activities) only" ~
      "Treated injury only",
    INJDVSTT == "Injury limiting activities and treated injury" ~
      "Limiting activities and treated injury")) %>% 
  
  # Covariate: Employment status
  mutate(employed = case_when(MAC_010 == "No" ~ "No",
                              MAC_010 == "Yes" ~ "Yes"))
```


## Dropping unwanted vars.
We only keep the redefined variables:
```{r}
cchs_withExclusions <- cchs_withExclusions %>% 
  select(pMentalHealth, physicalAct, age, sex, race,
         injury, employed)
```


## Making as factor
We convert all variables into factors to specify our desired reference levels.

```{r}
cchs_withExclusions <- cchs_withExclusions %>% 
  mutate(pMentalHealth = factor(pMentalHealth, levels = c("Fair-Poor",
                                                          "Excellent-Good"))) %>%
  mutate(physicalAct = factor(physicalAct,
                              levels = c("below act. level",
                                         "above act. level"))) %>%
  mutate(age = factor(age, levels = c("12-14", "15-17", "18-19",
                                      "20-24", "25-29", "30-34",
                                      "35-39", "40-44", "45-49",
                                      "50-54", "55-59", "60-64",
                                      "65-69", "70-74", "75-79",
                                      "80+"))) %>%
  mutate(sex = factor(sex, levels = c("Female", "Male"))) %>%
  mutate(race = factor(race, levels = c("White", "Non-white"))) %>%
  mutate(injury = factor(injury,
                         levels = c("No injuries",
                                    "Limiting activities only",
                                    "Treated injury only",
                                    "Limiting activities and treated injury"))) %>% 
  mutate(employed = factor(employed, levels = c("No", "Yes")))
```


## Inclusion/exclusion status
Only those respondents above 60 years old are included in our study.
Additionally, all observations with missing values for either the explanatory
variable (physical activity) or the outcome (perceived mental health) will be 
removed from the analytic sample.

### Coding included/excluded

```{r}
cchs_withExclusions <- cchs_withExclusions %>%
  mutate(inclusion = case_when((age %in% c("60-64", "65-69", "70-74",
                                          "75-79", "80+") &
                                  !is.na(physicalAct) &
                                  !is.na(pMentalHealth)) ~ "included",
                               TRUE ~ "excluded"))

table(cchs_withExclusions$inclusion, useNA = "ifany")
```


### Included vs. excluded

Included and excluded groups were not different with regard to sex (included: 
53.9% female; excluded: 53.6% female; p = 0.51); however, excluded respondents 
were more likely to be of non-white racial background (15.3% vs. 5.6%, 
p < 0.001), employed (77.9% vs. 39.3%, p < 0.001), and have activity-limiting 
injuries (15.1% vs. 10.5%, p < 0.001).

\

#### Sex
```{r}
test_result <- chisq.test(table("Inclusion" = cchs_withExclusions$inclusion,
                                "Sex" = cchs_withExclusions$sex))
test_result
test_result$expected
prop.table(test_result$observed, 1)
```

\

#### Race
```{r}
test_result <- chisq.test(table("Inclusion" = cchs_withExclusions$inclusion,
                                "Race" = cchs_withExclusions$race))
test_result
test_result$expected
prop.table(test_result$observed, 1)
```

\

#### Injury
```{r}
temp <- cchs_withExclusions %>% 
  mutate(injury = case_when(
    injury %in% c("Limiting activities only",
                  "Limiting activities and treated injury") ~ "Limiting act.",
    injury %in% c("No injuries", "Treated injury only") ~ "Not limiting act."))

test_result <- chisq.test(table("Inclusion" = temp$inclusion,
                                "Race" = temp$injury))
test_result
test_result$expected
prop.table(test_result$observed, 1)
```

\

#### Employment status
```{r}
test_result <- chisq.test(table("Inclusion" = cchs_withExclusions$inclusion,
                                "Race" = cchs_withExclusions$employed))
test_result
test_result$expected
prop.table(test_result$observed, 1)
```


## Applying the inclusion criteria
```{r}
cchs <- cchs_withExclusions %>% 
  filter(inclusion == "included") %>% 
  droplevels() %>%    # dropping extra levels
  select(-inclusion)  # dropping extra variables
```


## Level counts
The number of respondents in each level of our variable is as below:

```{r}
table(cchs$pMentalHealth, useNA = "ifany")     # Outcome: Perceived mental health
table(cchs$physicalAct, useNA = "ifany")       # Explanatory var: Physical activity
table(cchs$age, useNA = "ifany")               # Age
table(cchs$sex, useNA = "ifany")               # Sex
table(cchs$race, useNA = "ifany")              # Race
table(cchs$injury, useNA = "ifany")            # Injuries 
table(cchs$employed, useNA = "ifany")          # Employment status
```

\

***
# STUDY POPULATION

Of the `113,290` observations collected in the 2017-2018 CCHS, the final analytic 
sample included `25,806 (22.7%)` observations from respondents over 60 years old 
who provided valid responses to explanatory and outcome variables. Excluded 
observations consisted of 70,715 (62.4%) observations from respondents under 
60 years old, 16,030 (14.1%) invalid observations for whether a respondent met 
CPAG recommendations, which include participants who did not report physical 
activity minutes, and 739 invalid observations (0.7%) for perceived mental 
health status.

```{r}
cchs_exclusionType <- cchs_withExclusions %>%
  mutate(inclusionType = case_when(age %in% c("60-64", "65-69", "70-74",
                                              "75-79", "80+") ~ "included",
                                   TRUE ~ "excluded_age"))

cchs_exclusionType <- cchs_exclusionType %>%
  mutate(inclusionType = case_when(inclusionType == "excluded_age" ~ "excluded_age",
                                   !is.na(physicalAct) ~ "included",
                                   TRUE ~ "excluded_physicalAct"))

cchs_exclusionType <- cchs_exclusionType %>%
  mutate(inclusionType = case_when(inclusionType == "excluded_age" ~ "excluded_age",
                                   inclusionType == "excluded_physicalAct" ~ 
                                                      "excluded_physicalAct",
                                   !is.na(pMentalHealth) ~ "included",
                                   TRUE ~ "excluded_pMentalHealth"))

cchs_exclusionType %>% 
  count(inclusionType) %>% 
  mutate(Percentage = round(n / sum(n) * 100, 1))
```

\

Within the analytic sample, n = 9,156 (35.5%) respondents did not meet CPAG 
recommendations, while n = 16,650 (64.5%) were physically active at or above 
CPAG levels:

```{r}
cchs %>% 
  count(physicalAct) %>% 
  mutate(Percentage = round(n / sum(n) * 100, 1))
```

\

After re-coding CCHS perceived mental health answer options into 
two categories (Excellent-Good, including “excellent”, “very good”, and “good”, 
and Fair-Poor, including “fair” and “poor”), 24,456 (94.8%) respondents 
reported Excellent-Good and 1,350 (5.2%) respondents reported Fair-Poor mental 
health:
```{r}
cchs %>% 
  count(pMentalHealth) %>% 
  mutate(Percentage = round(n / sum(n) * 100, 1))
```

\

The Fair-Poor perceived mental health group comprised 6.30% of those 
that did not meet CPAG recommendations, and 4.64% of those who participated 
in physical activity at or above recommended CPAG levels:
```{r}
round(prop.table(table("Physical Activity" = cchs$physicalAct,
                       "Perceived Mental Health" = cchs$pMentalHealth,
                       useNA = "ifany"),
                 1) *100, 2)
```


\

***
# TABLE ONE
**Comparing Respondents who did and did not meet CPAG Recommendations:**  

Respondents who were physically active below CPAG recommendations were older 
(see **Table 1** below; p < 0.001), less likely to be male (40.0% vs. 49.4% male; 
p < 0.001), less likely to be injured (p < 0.001), and less likely to be 
employed (36.7% vs. 40.6%; p < 0.001) than were respondents who were physically 
active at or above CPAG recommendations. Respondents were similar in terms of 
racial background (5.5% vs. 5.7% non-white, including Aboriginal or other 
visible minority racial background; p = 0.509).

```{r, message=FALSE, warning=FALSE}
table_one_object <- CreateTableOne(vars = c("age", "sex", "race",
                                            "injury", "employed"),
                                   strata = "physicalAct",
                                   data = cchs,
                                   testApprox = chisq.test,
                                   argsApprox = list(correct = TRUE),
                                   testExact = fisher.test,
                                   testNormal = oneway.test,
                                   argsNormal = list(var.equal = FALSE),
                                   testNonNormal = kruskal.test)

table_one_printed <- print(table_one_object,
                           catDigits = 1,
                           contDigits = 1,
                           pDigits = 3,
                           # Supress printing in console:
                           printToggle = F)

# Creating a dataframe of the printed table one:
table_one <- as.data.frame(table_one_printed) %>%
  add_rownames(var = "Variable") %>%
  select(-test)
```


```{r}
table_one %>%
  mutate(Variable = cell_spec(Variable,
                              italic = ifelse(row_number() %in% c(3:7, 11:14),
                                              T, F))) %>%
  kable(escape = FALSE,
        caption = paste(c("Table 1: Participant characteristics: Canadian",
                          "Community Health Survey participants aged 60",
                          "years and above who did and did not participate",
                          "in physical activity as recommended by the",
                          "Canadian Physical Activity Guidelines (CPAG)."),
                        collapse = "")) %>%
  add_indent(c(3:7, 11:14)) %>% 
  kable_paper("hover", full_width = TRUE)
```


\

***
# (UN)ADJUSTED ANALYSES

Here are our unadjusted and adjusted models based on complete cases.
The unadjusted odds of a respondent who was physically active at or above 
recommended CPAG levels reporting Excellent-Good mental health were 1.38 
(95% confidence interval, CI: 1.24, 1.54) times that of a respondent who 
was physically active below recommended CPAG levels. After adjusting for 
age, sex, racial background, injury, and employment status, the odds ratio 
(OR) increased to 1.56 (95% CI: 1.37, 1.78).


## Bivariable Analysis
```{r}
model <- glm(pMentalHealth ~ physicalAct,
             family = binomial(link = logit),
             data = cchs)
publish(model)
```


## Multivariable Analysis
```{r}
model <- glm(pMentalHealth ~ physicalAct + age + sex + race +
                              injury + employed,
             family = binomial(link = logit), data = cchs)
publish(model)
```

\

***
# EVALUAT. MISSINGNESS

## Amount of missingness
The number of missing values and the percentage of the dataset missing for 
each variable is:

```{r, message=FALSE, warning=FALSE}
missing_data <- data.frame(n_missing = colSums(is.na(cchs)),
                           nrow = nrow(cchs)) %>%
  add_rownames() %>%
  mutate(pct_missing = round((n_missing/nrow)*100, 1))

missing_data
```


## Dummy-coded missingness
We added the missing values as another factor level to the dataset:

```{r}
cchs_missing <- cchs %>%
  mutate_if(is.factor,
            fct_explicit_na,
            na_level = "Missing")
```

\

An example to show what happened after dummy coding was applied:
```{r}
table(cchs$race, useNA="ifany")
table(cchs_missing$race, useNA="ifany")
```


## Sensitivity analysis
As confounding variables racial background (4.2% missing), injury 
(0.1% missing), and employment status (22.9% missing) contained missingness, 
a sensitivity analysis was performed to assess how the relationship between 
physical activity and perceived mental health status would change once 
missingness was included in the model with dummy coding:

```{r}
model <- glm(pMentalHealth ~ physicalAct + age + sex + race +
                              injury + employed,
             family = binomial(link = logit), data = cchs_missing)
publish(model)
```

Adjusting for age category, sex, racial background, injury, and employment 
status, a respondent who met CPAG recommendations was found to have 1.42 
(95% CI: 1.27, 1.60) times the odds of reporting having Excellent-Good mental 
health status than those of a respondent who did not meet CPAG recommendations. 
The calculated effect measure stayed almost the same after considering 
missingness. Therefore, we continued with complete case analysis.

\

***
# EFFECT MODIFICATION

Employment status was assessed for effect modification. If respondents were 
employed and met CPAG recommendations, they had 1.09 (95% CI: 0.85, 1.40) 
times the adjusted odds of reporting Excellent-Good mental health than those 
who did not meet recommendations. For those who were unemployed and who met 
CPAG recommendations, the OR was 1.81 (95% CI: 1.55, 2.12).

```{r}
data_employedYes <- cchs %>%
  filter(employed == "Yes")

data_employedNo <- cchs %>%
  filter(employed == "No")


model_employedYes <- glm(pMentalHealth ~ physicalAct + age + sex + race +
                                          injury,
                         family = binomial(link = logit),
                         data = data_employedYes)
model_employedNo <- glm(pMentalHealth ~ physicalAct + age + sex + race +
                                          injury,
                        family = binomial(link = logit),
                        data = data_employedNo)
```

For those employed:
```{r}
publish(model_employedYes)
```

For those unemployed:
```{r}
publish(model_employedNo)
```

\

***
# FOREST PLOT

Odds ratios of the five logistic regression models resulting from exploration 
for missingness and effect modification are depicted in the figure below:

```{r}
label <- c("Unadjusted model (A)",
           "Adjusted - Complete case analysis (B)",
           "Adjusted - Dummy-coded missingness (C)",
           "EM: Employed (D)",
           "EM: Unemployed (E)")
mean  <- c(1.38, 1.56, 1.42, 1.09, 1.81) 
lower <- c(1.24, 1.37, 1.27, 0.85, 1.55)
upper <- c(1.54, 1.78, 1.60, 1.40, 2.12)

df <- data.frame(label, mean, lower, upper)

# reverses the factor level ordering for labels after coord_flip()
df$label <- factor(df$label, levels=rev(df$label))

fp <- ggplot(data=df, aes(x=label, y=mean, ymin=lower, ymax=upper)) +
  geom_pointrange() + 
  geom_text(aes(label=mean),
            position = position_dodge(width = .9),
            vjust = -1,   
            size = 3) +
  geom_hline(yintercept=1, lty=2) +  # add a dotted line at x=1 after flip
  coord_flip() +                     # flip coordinates (puts labels on y axis)
  xlab("Logistic regression models") + ylab("Odds ratio (95% CI)") +
  theme_bw() 

print(fp)
```

\

***
\
\

<center>**The end!**</center>

